Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-24 ind_files/03_BD_processes.tex
Make changes as suggested by reviewers

I assume that the birth and death rate functions have the functional form
\begin{equation}
\label{nD_functional_forms_for_replicator}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= x_ib^{\textrm{(ind)}}_{i}(\mathbf{x}) + \mu Q_i(\mathbf{x})\\
d^{(K)}_i(\mathbf{x}) &= x_id^{\textrm{(ind)}}_i(\mathbf{x})
\end{aligned}
\end{equation}
where $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ and $d^{\textrm{(ind)}}_{i}(\mathbf{x})$ are non-negative functions that respectively describe the per-capita birth and death rate of type $i$ individuals, $\mu \geq 0$ is a constant describing the mutation rate in the population, and $Q_i(\mathbf{x})$ is a non-negative function that describes the additional birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$ that cannot be captured in the per-capita birth rate\footnote{When $x_i = 0$, \emph{i.e.} there are no type $i$ individuals in the population, individuals of type $i$ may still be born through mutations during births of the other types. This cannot be captured in $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ because the term $x_ib^{\textrm{(ind)}}_{i}(\mathbf{x})$ vanishes when $x_i = 0$. Note that no analogous problem exists for the death rate, since the death rate of type $i$ individuals must be 0 when $x_i$ is 0 to ensure that we never have negative population densities.}. My assumptions on the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates for each type. Let us define the \emph{Malthusian fitness} of the $i\textsuperscript{th}$ type as $w_i(\mathbf{x}) \coloneqq b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}(\mathbf{x})$, and the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type as $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + d^{\textrm{(ind)}}_{i}(\mathbf{x})$.  The quantity $w_i(\mathbf{x})$ describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$ discounting mutation. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but I will stick to $w_i$ and `fitness' here. $\tau_i$ is a measure of the number of events (birth events + death events) that a type $i$ individual experiences in a given time interval --- \textcolor{red}{\st{individuals}} \textcolor{blue}{Populations of types} with higher turnover rates experience more events (on average) than those with lower turnover rates. This can be thought of as a measure of the `pace of life' \textcolor{blue}{of a type}. I briefly note that the quantity $\tau_i$ has also been called the `variability in the reproductive output' in the literature~\citep{gillespie_natural_1974}. It is notable that both $w_i$ and $\tau_i$ depend on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and not just on the density of the focal type. Thus, in general, the fitness and the turnover rate in our model are both frequency and density-dependent.

@@

It is important to recognize that these quantities are distinct from and independent of the \emph{probabilistic} expectation, variance, and covariance obtained by integrating over realizations in the underlying probability space. Indeed, for finite populations, the statistical mean, statistical variance, and statistical covariance are all themselves stochastic processes: For each instant of time, these population-level quantities are a random variable (\emph{i.e.} a \emph{function} and not just a number) depending on $\mathbf{p}$, the (random) vector of type frequencies in the population. For infinite populations, the statistical mean, variance, and covariance are entirely deterministic time-dependent quantities that simply describe how $f$ is distributed across the population. \textcolor{blue}{Previous authors, such as Frank~\citep{frank_price_1997,frank_natural_2012} and van Veelen~\citep{van_veelen_use_2005,van_veelen_problem_2020}, have pointed out that} failure to clearly make this distinction between statistical operations and probabilistic operations has led to much confusion \textcolor{blue}{in the population biology literature} with regard to the infinite population Price equation, which is entirely deterministic and does not incorporate features of finite populations such as drift in its original formulation\textcolor{red}{\st{~(Frank, 1997; Van Veelen, 2005; Frank, 2012)}}.

@@

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' \textcolor{blue}{(Also see equation} \eqref{nD_stoch_RM_noise_term_alt_representation} \textcolor{blue}{in Appendix} \ref{App_alternative_representation} \textcolor{blue}{for a more elegant representation of these noise terms as an integral against a single Wiener process)}. \textcolor{blue}{ For the case $m=2, \mu = 0$ (two interacting types of individuals with no mutations between types), after using the representation of noise terms presented in Appendix \ref{App_alternative_representation} and letting $p = p_1$, equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation}
In this equation, if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. To see this, one can observe that the RHS of equation \eqref{two_species_stoch_RM} identically vanishes at both $p=0$ and $p=1$, since every term on the RHS contains the product $p(1-p)$. Further, since there is no mutation between types, $p=0$ (extinction of type 1 organisms) and $p=1$ (extinction of type 2 organisms) are both `absorbing states', \emph{i.e.} a system which reaches these states can never leave them. Well-behavedness at the boundaries for the more general equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{can be checked directly, and has been carried out in Appendix} \ref{App_well_behaved}.

>>>>

I assume that the birth and death rate functions have the functional form
\begin{equation}
\label{nD_functional_forms_for_replicator}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= x_ib^{\textrm{(ind)}}_{i}(\mathbf{x}) + \mu Q_i(\mathbf{x})\\
d^{(K)}_i(\mathbf{x}) &= x_id^{\textrm{(ind)}}_i(\mathbf{x})
\end{aligned}
\end{equation}
where $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ and $d^{\textrm{(ind)}}_{i}(\mathbf{x})$ are non-negative functions that respectively describe the per-capita birth and death rate of type $i$ individuals, $\mu \geq 0$ is a constant describing the mutation rate in the population, and $Q_i(\mathbf{x})$ is a non-negative function that describes the additional birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$ that cannot be captured in the per-capita birth rate\footnote{When $x_i = 0$, \emph{i.e.} there are no type $i$ individuals in the population, individuals of type $i$ may still be born through mutations during births of the other types. This cannot be captured in $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ because the term $x_ib^{\textrm{(ind)}}_{i}(\mathbf{x})$ vanishes when $x_i = 0$. Note that no analogous problem exists for the death rate, since the death rate of type $i$ individuals must be 0 when $x_i$ is 0 to ensure that we never have negative population densities.}. My assumptions on the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates for each type. Let us define the \emph{Malthusian fitness} of the $i\textsuperscript{th}$ type as $w_i(\mathbf{x}) \coloneqq b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}(\mathbf{x})$, and the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type as $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + d^{\textrm{(ind)}}_{i}(\mathbf{x})$.  The quantity $w_i(\mathbf{x})$ describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$ discounting mutation. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but I will stick to $w_i$ and `fitness' here. $\tau_i$ is a measure of the number of events (birth events + death events) that a type $i$ individual experiences in a given time interval ---  populations of types with higher turnover rates experience more events (on average) than those with lower turnover rates. This can be thought of as a measure of the `pace of life' of a type. I briefly note that the quantity $\tau_i$ has also been called the `variability in the reproductive output' in the literature~\citep{gillespie_natural_1974}. It is notable that both $w_i$ and $\tau_i$ depend on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and not just on the density of the focal type. Thus, in general, the fitness and the turnover rate in our model are both frequency and density-dependent.

@@

It is important to recognize that these quantities are distinct from and independent of the \emph{probabilistic} expectation, variance, and covariance obtained by integrating over realizations in the underlying probability space. Indeed, for finite populations, the statistical mean, statistical variance, and statistical covariance are all themselves stochastic processes: For each instant of time, these population-level quantities are a random variable (\emph{i.e.} a \emph{function} and not just a number) depending on $\mathbf{p}$, the (random) vector of type frequencies in the population. For infinite populations, the statistical mean, variance, and covariance are entirely deterministic time-dependent quantities that simply describe how $f$ is distributed across the population. Previous authors, such as Frank~\citep{frank_price_1997,frank_natural_2012} and van Veelen~\citep{van_veelen_use_2005,van_veelen_problem_2020}, have pointed out that failure to clearly make this distinction between statistical operations and probabilistic operations has led to much confusion in the population biology literature with regard to the infinite population Price equation, which is entirely deterministic and does not incorporate features of finite populations such as drift in its original formulation.

@@

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' (Also see equation \eqref{nD_stoch_RM_noise_term_alt_representation} in Appendix \ref{App_alternative_representation} for a more elegant representation of these noise terms as an integral against a single Wiener process). For the case $m=2, \mu = 0$ (two interacting types of individuals with no mutations between types), after using the representation of noise terms presented in Appendix \ref{App_alternative_representation} and letting $p = p_1$, equation \eqref{nD_eqn_for_frequencies} becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation}
In this equation, if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. To see this, one can observe that the RHS of equation \eqref{two_species_stoch_RM} identically vanishes at both $p=0$ and $p=1$, since every term on the RHS contains the product $p(1-p)$. Further, since there is no mutation between types, $p=0$ (extinction of type 1 organisms) and $p=1$ (extinction of type 2 organisms) are both `absorbing states', \emph{i.e.} a system which reaches these states can never leave them. Well-behavedness at the boundaries for the more general equation \eqref{nD_eqn_for_frequencies} can be checked directly, and has been carried out in Appendix \ref{App_well_behaved}.

### 2023-04-24 ind_files/05_unification_discussion.tex
Make changes as suggested by reviewers

So far, we have seen a lot of relatively abstract formalism for describing evolution in finite populations. Some concrete and hopefully illustrative examples are also presented in Appendix \ref{App_examples}. Have we gained anything from this (re)formulation? I show below that the \textcolor{red}{\st{formalism developed}} \textcolor{blue}{stochastic equation for trait frequencies I derived} in part \ref{part_theory} naturally yields some `fundamental equations' for evolutionary population dynamics that are very general, help us clearly understand how evolution operates in finite fluctuating populations, and recover well-known results in the infinite population limit, thus showing that I have generalized these infinite population results.  We will also see rigorous, quantitative formulations of some well-known ideas in evolutionary biology, such as mutation as a `source' of variation or drift causing continual erosion of variation.

@@

Equation \eqref{nD_eqn_for_frequencies}, which we derived in chapter \ref{chap_BD}, is a very general equation for how frequencies change over time in stochastic populations. To recap, we started with a population which can contain up to $m$ different (fixed) types of individuals. The population as a whole is characterized by a vector $\mathbf{n}(t) = [n_1(t), \ldots, n_m(t)]^{\mathrm{T}}$ indexing the number of individuals of each type in the population. Changes of the population are through either birth or death of single individuals. On ecological grounds, we postulated the existence of a `system-size' parameter $K>0$ that leads to density-dependent growth and prevents the population from growing infinitely large. We then moved from numbers $\mathbf{n}(t)$ to population densities $\mathbf{x}(t)=\mathbf{n}/K$ by assuming that birth and death rates depend on population densities $\mathbf{x}$ and not only on population numbers $\mathbf{n}$. In the regime where $K$ is not too small (corresponding to `medium sized' populations), we found a continuous description of how $\mathbf{x}$ changes stochastically. We further assumed that each type has a per-capita birth rate $b^{\textrm{(ind)}}(\mathbf{x})$, a per-capita death rate $d^{\textrm{(ind)}}(\mathbf{x})$, and an additional term $\mu Q_{i}(\mathbf{x})$ representing additional birth of type $i$ individuals due to mutations of other types. All three of these functions may in general vary in an arbitrarily complicated frequency-dependent manner (or be frequency independent constants, of course). We then moved from type densities $\mathbf{x}(t)$ to type frequencies $\mathbf{p}(t)$ and saw that \textcolor{blue}{under these assumptions,} $p_i$, the frequency of the $i\textsuperscript{th}$ type in the population, obeys:

@@

Each term in equation \eqref{nD_stochastic_Price} lends itself to a simple biological interpretation. The first term, $\textrm{Cov}(w,f)$, is well-understood in the classical Price equation and represents the effect of natural selection in the infinite population setting. In the stochastic Price equation \eqref{nD_stochastic_Price}, the effects of the second term of equation \eqref{nD_stochastic_RM} decompose into a selection term $\textrm{Cov}(\tau,f)$ for reduced turnover rates and a transmission bias term that vanishes in the weak mutation ($\mu \to 0$) limit. Following~\cite{constable_demographic_2016} and~\cite{week_white_2021}, I refer to the effect of the covariance term (the second term of equation \eqref{nD_stochastic_Price}) as \emph{noise-induced selection} since it occurs exactly analogously to classical natural selection (but for lower $\tau$) and is induced purely by the finiteness of the population. Biologically, the $\textrm{Cov}(\tau,f)$ term (with a negative sign) describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to gambler's ruin in probability theory through the following reasoning: If a type $i$ has a higher $\tau_i$, it must have a higher birth and death rate and thus experiences more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period and thus have greater stochastic fluctuations. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. \textcolor{blue}{Noise-induced selection for reduced turnover rates has also been interpreted as a selection for `lower reproductive variance'~\citep{gillespie_natural_1974,wang_reproductive_2023}. This latter interpretation is on purely mathematical grounds --- the total turnover $\tau_i x_i$ determines the variance of the distribution obtained as the solution to the SDE for population densities given by equation \eqref{nD_Ito_SDE}.}

\textcolor{red}{\st{Note that types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by selection (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavouring the beneficiary individual), which explains why this effect preferentially favors mutualists in reversing the direction of deterministic selection in finite population models with fluctuating population sizes if interaction effects are on the death rate~(McLeod and Day, 2019). Thus, selection for reduced turnover rate could help explain why cooperation often persists in fluctuating populations in laboratory experiments and finite population IbMs of social evolution that allow for fluctuating populations despite infinite population models predicting their extinction~(Houchmandzadeh and Vallade, 2012; Sanchez and Gore, 2013; Houchmandzadeh, 2015; Chotibut and Nelson, 2015; Behar et al., 2016; McAvoy et al., 2018; McLeod and Day, 2019)}}.

\textcolor{red}{\st{The fact that total population size controls the strength of noise-induced selection also explains why cooperation is favored in the early transient period of population growth~(Melbinger et al., 2010) when simulations are initiated from a small population size --- In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favoring cooperation. The fact that the entire term scales inversely with the total population size $KN_K(t)$ suggests that the effect of this force is weak for large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, timescale separation, or a weak selection + weak mutation limit~(McLeod and Day, 2019). In all three of these cases, the effects of classical selection vanish, and thus the first term on the RHS of }}\eqref{nD_stochastic_Price} \textcolor{red}{\st{becomes identically 0. It therefore no longer contributes to the trait frequency dynamics, allowing us to see the (otherwise weak) contributions of noise-induced selection.}}

@@

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by informally `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply informally taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} as the `fundamental equations' for the evolution of quantitative traits. In their supplementary sections S1 and S2,~\citet{week_white_2021} have recently proposed exactly the It\^o formula we would need for $L^2(\mathbb{R}, m)$ valued processes in which the $\mathcal{A}^{\pm}(x|\phi)$ (in my notation as defined in chapter \ref{chap_infD_processes}) are Hilbert-Schmidt operators via a heuristic It\^o multiplication table \textcolor{red}{\st{and}} \textcolor{blue}{. They have also} shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\citet{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details -the essence is that~\citet{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility and actually formulating the problem in formal terms rather than on mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied. The required properties will likely be fulfilled only for some `nice' subspace of $\mathcal{M}_K(\mathcal{T})$ and not the entire space, and future, more rigorous work should focus on figuring out which functions are included in this subspace.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to an SPDE describing populations with Gaussian trait distributions,~\citet{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} for the special case in which the type-level quantity $f$ is simply the value of the quantitative trait under study from a completely different starting point(!). Equation \eqref{disc_functional_field_eqns} is the `functional Fokker-Planck view' of the stochastic processes for the evolutionary ecology of quantitative traits that~\citet{week_white_2021} study from the `SPDE view'. Indeed, the two approaches are exactly complementary, and my formulation in chapter \ref{chap_infD_processes} provides an alternate method of attack for the study of quantitative traits in finite fluctuating populations that may be more appropriate for some particular problems, while~\posscite{week_white_2021} approach may be more appropriate for others.

>>>>

So far, we have seen a lot of relatively abstract formalism for describing evolution in finite populations. Some concrete and hopefully illustrative examples are also presented in Appendix \ref{App_examples}. Have we gained anything from this (re)formulation? I show below that the stochastic equation for trait frequencies I derived in part \ref{part_theory} naturally yields some `fundamental equations' for evolutionary population dynamics that are very general, help us clearly understand how evolution operates in finite fluctuating populations, and recover well-known results in the infinite population limit, thus showing that I have generalized these infinite population results.  We will also see rigorous, quantitative formulations of some well-known ideas in evolutionary biology, such as mutation as a `source' of variation or drift causing continual erosion of variation.

@@

Equation \eqref{nD_eqn_for_frequencies}, which we derived in chapter \ref{chap_BD}, is a very general equation for how frequencies change over time in stochastic populations. To recap, we started with a population which can contain up to $m$ different (fixed) types of individuals. The population as a whole is characterized by a vector $\mathbf{n}(t) = [n_1(t), \ldots, n_m(t)]^{\mathrm{T}}$ indexing the number of individuals of each type in the population. Changes of the population are through either birth or death of single individuals. On ecological grounds, we postulated the existence of a `system-size' parameter $K>0$ that leads to density-dependent growth and prevents the population from growing infinitely large. We then moved from numbers $\mathbf{n}(t)$ to population densities $\mathbf{x}(t)=\mathbf{n}/K$ by assuming that birth and death rates depend on population densities $\mathbf{x}$ and not only on population numbers $\mathbf{n}$. In the regime where $K$ is not too small (corresponding to `medium sized' populations), we found a continuous description of how $\mathbf{x}$ changes stochastically. We further assumed that each type has a per-capita birth rate $b^{\textrm{(ind)}}(\mathbf{x})$, a per-capita death rate $d^{\textrm{(ind)}}(\mathbf{x})$, and an additional term $\mu Q_{i}(\mathbf{x})$ representing additional birth of type $i$ individuals due to mutations of other types. All three of these functions may in general vary in an arbitrarily complicated frequency-dependent manner (or be frequency independent constants, of course). We then moved from type densities $\mathbf{x}(t)$ to type frequencies $\mathbf{p}(t)$ and saw that under these assumptions, $p_i$, the frequency of the $i\textsuperscript{th}$ type in the population, obeys:

@@

Each term in equation \eqref{nD_stochastic_Price} lends itself to a simple biological interpretation. The first term, $\textrm{Cov}(w,f)$, is well-understood in the classical Price equation and represents the effect of natural selection in the infinite population setting. In the stochastic Price equation \eqref{nD_stochastic_Price}, the effects of the second term of equation \eqref{nD_stochastic_RM} decompose into a selection term $\textrm{Cov}(\tau,f)$ for reduced turnover rates and a transmission bias term that vanishes in the weak mutation ($\mu \to 0$) limit. Following~\cite{constable_demographic_2016} and~\cite{week_white_2021}, I refer to the effect of the covariance term (the second term of equation \eqref{nD_stochastic_Price}) as \emph{noise-induced selection} since it occurs exactly analogously to classical natural selection (but for lower $\tau$) and is induced purely by the finiteness of the population. Biologically, the $\textrm{Cov}(\tau,f)$ term (with a negative sign) describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to gambler's ruin in probability theory through the following reasoning: If a type $i$ has a higher $\tau_i$, it must have a higher birth and death rate and thus experiences more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period and thus have greater stochastic fluctuations. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. Noise-induced selection for reduced turnover rates has also been interpreted as a selection for `lower reproductive variance'~\citep{gillespie_natural_1974,wang_reproductive_2023}. This latter interpretation is on purely mathematical grounds --- the total turnover $\tau_i x_i$ determines the variance of the distribution obtained as the solution to the SDE for population densities given by equation \eqref{nD_Ito_SDE}.

@@

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by informally `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply informally taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} as the `fundamental equations' for the evolution of quantitative traits. In their supplementary sections S1 and S2,~\citet{week_white_2021} have recently proposed exactly the It\^o formula we would need for $L^2(\mathbb{R}, m)$ valued processes in which the $\mathcal{A}^{\pm}(x|\phi)$ (in my notation as defined in chapter \ref{chap_infD_processes}) are Hilbert-Schmidt operators via a heuristic It\^o multiplication table. They have also shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\citet{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details -the essence is that~\citet{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility and actually formulating the problem in formal terms rather than on mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied. The required properties will likely be fulfilled only for some `nice' subspace of $\mathcal{M}_K(\mathcal{T})$ and not the entire space, and future, more rigorous work should focus on figuring out which functions are included in this subspace.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to an SPDE describing populations with Gaussian trait distributions,~\citet{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} for the special case in which the type-level quantity $f$ is simply the value of the quantitative trait under study from a completely different starting point(!). Equation \eqref{disc_functional_field_eqns} is the `functional Fokker-Planck view' of the stochastic processes for the evolutionary ecology of quantitative traits that~\citet{week_white_2021} study from the `SPDE view'. Indeed, the two approaches are exactly complementary, and my formulation in chapter \ref{chap_infD_processes} provides an alternate method of attack for the study of quantitative traits in finite fluctuating populations that may be more appropriate for some particular problems, while~\posscite{week_white_2021} approach may be more appropriate for others.
